Determinism, Complexity, and Predictability in Computer Performance

### 2014-07-29 wpe.tex
working through

From a different point of view, we can consider the information (as
measured by the Shannon entropy) contained in a single observable, say
the \emph{present}, of the system. This information can be partitioned
into two components: the information shared with past
observations---e.g., the mutual information between the past and
present---and the information in the present that is not contained in
the past (aka ``the conditional entropy of the present given the
past'').  The first part is known as the \emph{redundancy}, as it is
information in the present that is also in the past.  The second part
is the aforementioned \emph{Shannon entropy rate}.  It seems obvious
that the more redundancy in a signal, the more predictable it is.
Redundancy is critical to prediction, in general.  And the specific
\emph{form} of the redundancy dictates whether a particular prediction
method will work well or poorly on the corresponding signal.  A linear
method cannot capture or make use of nonlinear redundancy, for
instance.  This issue---which has been explored extensively using
clean data and/or systems where the generating process is known, but
not so much with empirical data from unknown systems---is central to
the claims in this paper and the discussion in the following section.

Previous approaches to measuring temporal complexity via the Shannon
entropy rate \cite{Shannon1951, mantegna1994linguistic} required
categorical data: $x_i \in \mathcal{S}$ for some finite or countably
infinite \emph{alphabet} $\mathcal{S}$.  Data taken from real-world
systems are, however, effectively\footnote{Measurements from
  finite-precision sensors are discrete, but data from modern
  high-resolution sensors are, for the purposes of entropy
  calculations, effectively continuous.}  real-valued.  To analyze
real-valued data using a method that requires categorical values, one
must discretize the data---typically by binning.  Unfortunately, this
is rarely a good solution, as the binning of the values introduces
spurious dynamics~\cite{bollt2001}.  The field of symbolic dynamics
offers discretization methods that do not disturb the intrinsic
behavior.  These methods are, however, fragile in the face of noise;
worse yet, they require knowledge of the underlying system.  This is
not useful in our context, where the target of study is experimental
time-series data.

>>>>

From a different point of view, we can consider the information (as
measured by the Shannon entropy) contained in a single observable, say
the \emph{present}, of the system. This information can be partitioned
into two components: the information shared with past
observations---e.g., the mutual information between the past and
present---and the information in the present that is not contained in
the past (aka ``the conditional entropy of the present given the
past'').  The first part is known as the \emph{redundancy}, as it is
information in the present that is also in the past.  The second part
is the aforementioned Shannon entropy rate.  It seems obvious that the
more redundancy in a signal, the more predictable it should be.
Redundancy is critical to prediction, in general.  And the specific
\emph{form} of the redundancy should dictate whether a particular
prediction method will work well or poorly on the corresponding
signal.  A linear method cannot capture or make use of nonlinear
redundancy, for instance.  This issue---which has been explored
extensively using clean data and/or systems where the generating
process is known, but not so much with empirical data from unknown
systems---is central to the claims in this paper and the discussion in
the following section.

Previous approaches to measuring temporal complexity via the Shannon
entropy rate \cite{Shannon1951, mantegna1994linguistic} required
categorical data: $x_i \in \mathcal{S}$ for some finite or countably
infinite \emph{alphabet} $\mathcal{S}$.  Data taken from real-world
systems are, however, effectively\footnote{Measurements from
  finite-precision sensors are discrete, but data from modern
  high-resolution sensors are, for the purposes of entropy
  calculations, effectively continuous.}  real-valued.  To analyze
real-valued data using a method that requires categorical values, one
must discretize the data---typically by binning.  Unfortunately, this
is rarely a good solution, as the binning of the values introduces
spurious dynamics~\cite{bollt2001}.  The field of symbolic dynamics
offers discretization methods that do not disturb the intrinsic
behavior.  These methods are, however, fragile in the face of noise;
worse yet, they require knowledge of the underlying system.  This is
inappropriate in our context, where the target of study is
experimental time-series data.
